---
title: "Getting Started with OHLviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with OHLviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview

OHLviz creates consistent, customizable charts for Ontario Hockey League
data. Its plotting functions accept data frames and return ggplot objects.

OHLpkg handles data retrieval, preparation, filtering, and analysis.
OHLviz handles visualization. You can also use data from other sources
by specifying column names.

```{r setup}
library(OHLviz)
```

## Prepare your data

Supply one season and a consistent reporting scope per chart.
Apply qualification filters before plotting, and resolve duplicate
records or player team splits.

The following fictional data uses OHLpkg's column names:

```{r skater-data}
skaters <- data.frame(
  Name = c("Player A", "Player B", "Player C", "Player D", "Player E"),
  Team = c("North", "North", "North", "South", "South"),
  GP = c(60, 58, 61, 59, 60),
  G = c(35, 28, 31, 40, 22),
  A = c(52, 48, 39, 42, 44)
)

skaters$PTS <- skaters$G + skaters$A
skaters[["Pts/G"]] <- skaters$PTS / skaters$GP
```

These calculations happen before plotting. OHLviz does not retrieve
data or automatically calculate a points-per-game column.

## Plot scoring leaders

By default, `plot_scoring_leaders()` ranks players by `PTS`.

```{r scoring-leaders}
plot_scoring_leaders(
  skaters,
  n = 5,
  title = "Scoring Leaders",
  subtitle = "Example player comparison",
  caption = "Illustrative data | OHLviz"
)
```

Change `stat` to select another non-negative numeric column:

```{r goals-leaders}
plot_scoring_leaders(
  skaters,
  stat = "G",
  n = 3,
  title = "Goal Leaders",
  caption = "Illustrative data | OHLviz"
)
```

Common choices include `"PTS"`, `"G"`, `"A"`, and `"Pts/G"`.

Players are ranked highest first. Ties preserve input row order, and
`n` limits the chart to that many rows. If fewer players are available,
the chart displays all available players.

For rate statistics, choose an appropriate qualification threshold
before plotting:

```{r scoring-rate}
qualified_skaters <- skaters[
  !is.na(skaters$GP) & skaters$GP >= 60,
  ,
  drop = FALSE
]

plot_scoring_leaders(
  qualified_skaters,
  stat = "Pts/G",
  title = "Points per Game",
  subtitle = "Minimum 60 games played",
  caption = "Illustrative data | OHLviz"
)
```

## Compare scoring within a team

`plot_team_scoring()` displays goals and assists as stacked bars.
Players are ranked by their combined total.

```{r team-scoring}
plot_team_scoring(
  skaters,
  team = "North",
  title = "North Team Scoring",
  caption = "Illustrative data | OHLviz"
)
```

The total at the end of each bar is calculated as goals plus assists.
The function does not require a separate `PTS` column.

If your data contains exactly one named team, you can omit `team`:

```{r single-team, eval=FALSE}
north_skaters <- skaters[skaters$Team == "North", ]

plot_team_scoring(north_skaters)
```

Selecting a team filters rows. It does not separate a traded player's
season totals into statistics earned with each team. Supply team-specific
statistics when that distinction matters.

Duplicate player names within the selected team produce an error.
Resolve duplicate records, or provide distinct labels when two different
players share a name.

## Compare goalies

Create a fictional goalie dataset:

```{r goalie-data}
goalies <- data.frame(
  Name = c("Goalie A", "Goalie B", "Goalie C", "Goalie D"),
  Team = c("North", "South", "East", "West"),
  GP = c(42, 45, 38, 40),
  GAA = c(2.45, 2.70, 2.30, 3.05),
  W = c(25, 30, 22, 18)
)

goalies[["SAV%"]] <- c(0.915, 0.908, 0.921, 0.899)
```

The goalie function supports three metrics:

| `stat` | Metric | Ranking |
|---|---|---|
| `"SAV%"` | Save percentage, the default | Highest first |
| `"GAA"` | Goals-against average | Lowest first |
| `"W"` | Wins | Highest first |

```{r goalie-leaders}
plot_goalie_leaders(
  goalies,
  stat = "GAA",
  title = "Goals-Against Average Leaders",
  caption = "Illustrative data | OHLviz"
)
```

Goalie charts use dots. Their horizontal axes are fitted to the selected
values and do not necessarily start at zero.

Qualification filters belong upstream. These charts rank the supplied
goalies; they do not automatically apply official league qualification rules.

### Save percentage units

By default, save percentage inputs must be proportions such as `0.915`.

```{r save-percentage}
plot_goalie_leaders(
  goalies,
  stat = "SAV%",
  title = "Save Percentage Leaders",
  caption = "Illustrative data | OHLviz"
)
```

If your source uses values such as `91.5`, specify `"percent"` explicitly:

```{r percent-input, eval=FALSE}
percent_goalies <- goalies
percent_goalies[["SAV%"]] <- percent_goalies[["SAV%"]] * 100

plot_goalie_leaders(
  percent_goalies,
  save_pct_scale = "percent"
)
```

OHLviz does not guess the input units. Both formats are displayed as
proportions with three decimal places. GAA uses two decimal places,
and wins use whole numbers.

## Use different column names

OHLviz does not require data to come from OHLpkg.

For a scoring leaderboard, specify the name and statistic columns.
Use `team_col = NULL` when no team column is available:

```{r custom-columns}
custom_skaters <- data.frame(
  player = c("Alex", "Sam", "Jordan"),
  points = c(65, 72, 58)
)

plot_scoring_leaders(
  custom_skaters,
  stat = "points",
  player_col = "player",
  team_col = NULL,
  title = "Player Points",
  caption = "Illustrative data | OHLviz"
)
```

For goalie charts, `stat` identifies the metric and determines ranking
and formatting. `stat_col` identifies the column containing its values:

```{r custom-goalie-columns, eval=FALSE}
custom_goalies <- data.frame(
  goalie = c("Alex", "Sam"),
  average = c(2.80, 2.20)
)

plot_goalie_leaders(
  custom_goalies,
  stat = "GAA",
  stat_col = "average",
  player_col = "goalie",
  team_col = NULL
)
```

## Customize your charts

All plotting functions support `title`, `subtitle`, and `caption`.

Scoring and goalie leaderboards use `colour`. Team scoring charts use
`goals_colour` and `assists_colour`.

```{r colours, eval=FALSE}
plot_team_scoring(
  skaters,
  team = "North",
  goals_colour = "#142D4E",
  assists_colour = "#4FA3D1",
  caption = "Illustrative data | OHLviz"
)
```

Returned ggplot objects support additional customization:

```{r customize-ggplot, eval=FALSE}
p <- plot_scoring_leaders(skaters)

p +
  ggplot2::labs(title = "My Scoring Leaderboard") +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 18)
  )
```

### Use the theme independently

`theme_ohl()` can also style your own ggplot2 charts:

```{r standalone-theme}
ggplot2::ggplot(
  skaters,
  ggplot2::aes(x = G, y = A)
) +
  ggplot2::geom_point(
    colour = "#2474B5",
    size = 3
  ) +
  ggplot2::labs(
    title = "Goals and Assists",
    x = "Goals",
    y = "Assists",
    caption = "Illustrative data | OHLviz"
  ) +
  theme_ohl()
```

Use `base_size` and `base_family` to customize typography.
The default family is `"sans"` for portability.

The theme controls text, backgrounds, spacing, and gridlines.
Data colours are set separately through chart arguments or colour scales.

## Export a chart

Store the plot and save it with `ggplot2::ggsave()`:

```{r export, eval=FALSE}
p <- plot_scoring_leaders(
  skaters,
  caption = "Illustrative data | OHLviz"
)

ggplot2::ggsave(
  filename = "scoring-leaders.png",
  plot = p,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

For a PDF, change the filename extension to `.pdf`.

Increase chart height when displaying more players. Increase width
when player names, team names, or titles need additional room.
Inspect the exported chart at its intended display size.

## Handle input problems

The functions check required columns and numeric statistic types.

- Missing or non-finite scoring values are omitted with a warning.
- If no usable values remain, plotting stops with an error.
- Negative scoring values are rejected.
- Team goals, team assists, and goalie wins must be whole numbers.
- Missing or blank player names in otherwise usable rows are rejected.

Resolve data problems upstream rather than replacing missing statistics
with zero unless zero is the correct value.

## Connect to OHLpkg

With OHLpkg installed, retrieve and filter your data before plotting.

The following example is not executed when this guide builds because
it requires a live data connection:

```{r ohlpkg-workflow, eval=FALSE}
skater_data <- OHLpkg::get_Stats(
  season_name = "2026 Season",
  min_games = 10
)

plot_scoring_leaders(
  skater_data,
  n = 10,
  subtitle = "2026 Season | Minimum 10 games played",
  caption = "Data: OHLpkg | Visualization: OHLviz"
)

goalie_data <- OHLpkg::get_GoalieStats(
  season_name = "2026 Season",
  min_games = 10
)

plot_goalie_leaders(
  goalie_data,
  stat = "GAA",
  n = 10,
  subtitle = "2026 Season | Minimum 10 games | Lower is better",
  caption = "Data: OHLpkg | Visualization: OHLviz"
)
```

Check the units of save percentage in your source before plotting it.
OHLpkg is optional for using OHLviz; the plotting functions operate
directly on supplied data frames.

## Function help

For complete argument documentation and examples:

```{r help, eval=FALSE}
?plot_scoring_leaders
?plot_team_scoring
?plot_goalie_leaders
?theme_ohl
```
